---
title: "ESRM 6990V Class Project"
author: "Henrietta Kadi Tettey-Tawiah"
format: pdf
editor: visual
execute: 
  echo: false #hides code 
  message: false
  error: false
  warning: false
  #output: false
bibliography: references.bib
---

# TIMSS DATA

## Introduction: Research Scenario and Research Questions

The researcher would want to investigate the performance of students in the United States of America (USA) and Canada based on the 2023 [TIMSS](https://timss2023.org/data) (Trends in International Mathematics and Science Study) dataset. This analysis focuses on math performance of 4th grade students using their percentage correct on the TIMSS math achievement test. In this study, student level factors, home factors as well as teacher-level factors are explored to identify their respective impacts on an individual student's math test score. Specifically, this research;

1.  Explores how factors such as gender, country, student confidence among others affect the performance of a student on the math test.
    -   Explore if there are any country-level and/or sex differences in math percentage scores. ("How does math performance vary across countries and sexes?")
    -   Examines how various teacher characteristics influence student's score on the math test

## Data Analysis

### Student-Level Analysis

The dataset used in this contained a total of 18369 participants after data cleaning. The dataset had a total of 373 variables but only 13 are reported below:

-   **Student_ID:** Student Identifier

-   **Country:** Country of Participation (CAN = 1, USA = 2)

-   **Sex:** Sex/Gender (1 = Girl, 2 = Boy)

-   **Age :** Age of student

-   **Number_of_Home_Study_Supports:** Derived from subscale measuring resources available to the student at home.

-   **Disorderly_Behavior:** Scale score from the subscale measuring disorderly behavior during math lessons.

-   **Instructional_Clarity:** Scale score from the subscale measuring instructional clarity in mathematics lessons.

-   **Digital_Self_Efficacy:** Scale score from the subscale measuring digital self-efficacy.

-   **Sense_of_School_Belonging:** Scale score from the subscale measuring students sense of school belonging.

-   **Student_Bullying:** Scale score from the subscale measuring student bullying.

-   **Like_Learning_Math:** Scale score from the subscale measuring if students like learning mathematics.

-   **Confident_in_Math:** Scale score from the subscale measuring students confidence in mathematics.

-   **Math_Percent_Correct**: Students' percent correct on the 2023 TIMSS Math Achievement Test.

The scale scores reported were derived following several steps including; item parameter estimation, item fit evaluation among others [@yin2024].

#### Data Preparation

The data preparation stage began with renaming the columns in the `Home_Student` dataset with more descriptive names. After this, the dataset is cleaned by removing any rows containing missing values resulting in a final dataset with only complete cases as well as specifying categorical variables and dropping any unused levels.

#### Descriptive Statistics

The total sample size of the Canada and U.S.A. 4th grade students was 22806. However, due to non-response, the sample size used for this analysis was 18369; 9351 girls and 9018 boys. There were more Canadian 4th Grade students compared to the U.S.A. in the sample (@fig-Country-by-Sex). The overall mean age was 10 years (S.D. = 0.390) and the average sample class score was 52% (S.D. = 0.390). About 51% of the entire sample was female with Canada accounting for about 52% of the total population of females. A comprehensive summary of the sample information can be seen in @tbl-sample-descriptives. The distribution of the math scores for each country and sex subgroup can be described asbimodal as shown in @fig-dist-of-math-score. As shown in @fig-dist-of-math-score-country-sex-home-support, it can be observed that U.S.A students had a higher Math score average with boys also having a higher average score. Students who had access to both their own rooms and internet access seemed to have a better average score compared to those who had neither (@fig-dist-of-math-score-country-sex-home-support-4).

```{r}
#| label: tbl-sample-descriptives
#| tbl-cap: Sample Descriptive Statistics

source("C:/Users/Henretta Tawiah/Desktop/School/PhD/Spring 2025/R/Project/ESRM6990V/ESRM6990V_RScript.R")

Tbl1


```

```{r}
#| label: fig-Country-by-Sex
#| fig-cap: Number of sudents by Country and Sex

Fig1
```

```{r}
#| label: fig-dist-of-math-score
#| fig-cap: Histogram of the Distribution of Math Scores

Fig2
```

```{r}
#| label: fig-dist-of-math-score-country-sex-home-support
#| layout-ncol: 2
#| layout-nrow: 2
#| fig-cap: Boxplots of the the Distribution of Scores by Country, Sex and Home Support
#| fig-subcap:
#| - "Boxplots of the the Distribution of Scores by Country"
#| - "Boxplots of the the Distribution of Scores by Sex "
#| - "Boxplots of the the Distribution of Scores by Country and Sex"
#| - Boxplots of the the Distribution of Scores by Home Support
#

Fig3
Fig4
Fig5
Fig6
```

```{r}
#| label: fig-Correlations
#| fig-cap: Correlation Heatmap for Student-Level Analysis

Fig7
```

#### Regression Analysis

A step-wise regression model was estimated to evaluate whether the independent variables *(Country, Sex, Age, Disorderly_Behavior, Instructional_Clarity, Digital_Self_Efficacy, Student_Bullying, Like_Learning_Math, Confident_in_Math, and Number_of_Home_Study_Supports)* could predict the dependent variable, *Math_Percent_Correct* . The linear relationship between independent variables and math score was found to be statistically significant $F(11, 18357) = 439.39,\ p < .001$. As shown in @fig-Correlations, the correlations between the predictors and math score were in the moderate to low range. Appoximately $21\%$ of the total variation in math percent score was account for by its linear relationship with the predictors, $R^2 = .21,\ R^2_{adj}=.21$. Among the predictors, instructional clarity had a statistically non-significant and negative effect on math percentage value $(\beta = -.16,\ p < .001)$, suggesting that greater instructional clarity was not associated with higher math percent values in this model over and above all other predictors. The effect of Country (U.S.A.) was statistically significant and positive $(\beta = 5.58,\ p < .001)$, suggesting that holding all other predictors unchanged, 4th grade students scored 5.58% higher than those in Canada. The effect of Sex (boy) was also statistically significant and positive $(\beta = 2.01,\ p < .001)$, suggesting that controlling for all other predictors, boys in the 4th grade scored 2.01% higher than girls. Interestingly, Like_Learning_Math was found to have statistically significant and negative effect on math percent score $(\beta = -.97,\ p < .001)$. This suggests that partialling out the effects of all other predictors, and a unit increase in a student whether a student likes learning math results in a -0.97% decrease in math percent score. All other predictors in the model had a statistically significant and positive effect on math percent score. This suggests that holding all other predictirs unchanged, an increase in their respective values is expected lead to an increase in math percentage score. The details of the results from the step-wise analysis are presented in @tbl-Initial-regression-for-student-data. The predictor Confident_in_Math had the largest effect on the math percent score followed by Number_of_Home_Study_Supports_Either @tbl-Initial-regression-for-student-data-4. The predictors used in this analysis do not exhibit multicollinearity as shown in @tbl-Initial-regression-for-student-data-3. A look at the residual plots in @fig-residual-std_reg_initial show a slight non-linearity with possible non-normality from the Q-Q plot.

```{r}
#| label: tbl-Initial-regression-for-student-data
#| tbl-cap: Student-Level Regression Analysis
#| tbl-subcap: 
#|     - Student Level Regression Coefficients
#|     - Student Level Model Summary Statistics
#|     - Variance Inflation Factors
#|     - Standardized Regression Coefficients

Tbl2
Tbl3
Tbl4
Tbl5
```

```{r}
#| label: fig-residual-std_reg_initial
#| tbl-cap: Residual Plots for the First Regression Model

 par(mfrow=c(2, 2))
  plot(std_stepwise)
```

##### Difference-in-Difference Analysis on Combinations of Sex and Country

This section of the analysis focuses on whether there are sex and/or country specific differences in the performance of the 4th grade students on the TIMSS math test. To do this, a differences-in-differences analysis was performed on various combinations of the country and sex variables. From @tbl-Differences-in-Differences-tables and @fig-Differences-in-Differences-fiqures, it can be observed that holding all other predictors unchanged, the predicted average math percent value for a typical 4th grade student in the U.S.A is higher than in Canada. Also, a typical Canadian or U.S.A 4th grade boy seem to have a higher predicted average math percent score than their girl counterparts. A marginal contrast analysis (@tbl-Differences-in-Differences-tables-1) was further performed to evaluate if these differences were statistically significant. This analysis revealed that the differences in predicted scores across all combinations of sex and country were statistically significant. In the U.S.A, boys had significantly higher predicted math scores than girls $(\text{difference}= 2.01,\ 95\%  \text{ CI } [1.38, 2.63],\ t(18357) = 6.28,\ p < .001)$. Boys in Canada also had a significantly higher predicted math scores than girls$(\text{difference}= 2.01,\ 95\%  \text{ CI } [1.38, 2.63],\ t(18357) = 6.28,\ p < .001)$. When comparing the average predicted scores of girls in the U.S.A. to those in Canada, the girls in the U.S.A had a significantly higher score $(\text{difference}= 5.58,\ 95\%   \text{ CI } [4.91, 6.26],\ t(18357) = 16.12,\ p < .001)$. Similarly, girls in the U.S.A. also outperformed the boys in Canada $(\text{difference}= -3.58,\ 95\%   \text{ CI } [–4.51, –2.65],\ t(18357) = -7.53,\ p < .001)$. Additionally, boys in the U.S.A scored significantly higher than the boys in Canada $(\text{difference}= 5.58,\ 95\%   \text{ CI } [4.91, 6.26],\ t(18357) = 16.12,\ p < .001)$. In the same vein, boys in the U.S.A. also outperformed the girls in Canada by the largest margin $(\text{difference}= 7.59,\ 95\%   \text{ CI } [6.67, 8.51],\ t(18357) = 16.25,\ p < .001)$.

These findings suggest that while gender differences exist, the national level factors may play a more substantial role in the TIMSS math percent score of 4th grade students. Further research should explore some potential national level contributors like curriculum standards.

```{r}
#| label: tbl-Differences-in-Differences-tables
#| tbl-cap: Differences-in-Differences Analysis
#| tbl-subcap: 
#|      - Estimated Marginal Means of Predicted Math Scores by Sex and Country
#|      - Pairwise Contrasts of Predicted Math Scores by Sex and Country

Tbl6
Tbl7
```

```{r}
#| label: fig-Differences-in-Differences-fiqures
#| fig-cap: Estimated Marginal Means of Predicted Math Scores by Sex and Country

plot(std_pred)
```

### Teacher-Level Analysis

The dataset used in this contained a total of 18194 participants after data cleaning. The dataset had a total of 538 variables but only 14 are reported below:

-   **Teacher_ID:** Teacher Identifier

-   **Years_Teaching:** Years teacher has been teaching

-   **Sex:** Sex/Gender of teacher (1: Female; 2: Male; 3: Other)

-   **Age:** Age of Teacher (1: Under 25; 2: 25–29; 3: 30–39; 4: 40–49; 5: 50–59; 6: 60 or more)

-   **Level_of_Formal_Educ:** Level of formal education teacher has completed (1: Did not complete Upper secondary education—ISCED Level 3; 2: Upper secondary education—ISCED Level 3 (have not completed postsecondary or tertiary education); 3: Post-secondary, non-tertiary education—ISCED Level 4; 4: Short-cycle tertiary education—ISCED Level 5; 5: Bachelor’s or equivalent level—ISCED Level 6; 6: Master’s or equivalent level—ISCED Level 7; 7: Doctor or equivalent level—ISCED Level 8)

-   **Class_Size:** Number of students in the class.

-   **Homework_Freq:** How often math homework is assigned (1: I do not assign mathematics homework; 2: Less than once a week; 3: 1 or 2 times a week; 4: 3 or 4 times a week; 5: Every day)

-   **Tch_Academic_Success:** Scale score from the subscale measuring school emphasis on academic success-teacher.

-   **Safe_Orderly_Schools :** Scale score from the subscale measuring safe and orderly schools-teacher.

-   **Tch_Job_Satis :** Scale score from the subscale measuring teachers job satisfaction.

-   **Student_not_Ready:** Scale score from the subscale measuring teaching limited by student not ready.

-   **Tch_Educ_Math_Major :** Teachers majored in education and mathematics (1: Major in Edu and Math; 2: Major in Edu but not Math; 3: Major in Math but not Edu; 4: All other Majors; 5: No Formal Edu Beyond Upper Secondary).

-   **Math_Instruction_Hours:** Mathematics instruction hours per week

-   **Math_Percent_Correct**: Students' percent correct on the 2023 TIMSS Math Achievement Test.

The scale scores reported were derived following several steps including; item parameter estimation, item fit evaluation among others [@yin2024].

#### Data Preparation

As with the `Home_Student` dataset, the data preparation for the `Teacher_Student` dataset started with renaming the original variables with more descriptive names. After this, the dataset is cleaned by removing any rows containing missing values resulting in a final dataset with only complete cases as well as specifying categorical variables and dropping any unused levels.

```{r}
#| label: Cleaning Teacher Data


```

#### Descriptive Statistics

The total sample size of the Canada and U.S.A. 4th grade students was 22806. However, due to non-response, the sample size used for this analysis was 18369; 9351 girls and 9018 boys. There were more Canadian 4th Grade students compared to the U.S.A. in the sample (@fig-Country-by-Sex). The overall mean age was 10 years (S.D. = 0.390) and the average sample class score was 52% (S.D. = 0.390). About 51% of the entire sample was female with Canada accounting for about 52% of the total population of females. A comprehensive summary of the sample information can be seen in @tbl-sample-descriptives. The distribution of the math scores for each country and sex subgroup can be described asbimodal as shown in @fig-dist-of-math-score. As shown in @fig-dist-of-math-score-country-sex-home-support, it can be observed that U.S.A students had a higher Math score average with boys also having a higher average score. Students who had access to both their own rooms and internet access seemed to have a better average score compared to those who had neither (@fig-dist-of-math-score-country-sex-home-support-4).

```{r}
#| label: tbl-tch-sample-descriptives
#| tbl-cap: Sample Descriptive Statistics for Teacher Dataset

```
